---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:** Our group's mean proportion of colored M&Ms was far off from Rob's sample, which was about 53%. Ours was 92.45%. This is different because we took different samples from the total population, which may not have had equal distribution or fair sampling. If we combine all samples, the distributions will likely normalize and the extreme results will be erased and offset. In real-life polls and surveys, weighed or biased sampling methods can cause these unequal--and unreliable--sampling results. 

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:** The starkest difference between the population and the sample is the different in proportion of men in the sample. Men, which make up about 50% of campus, only represented 38% of the respondents. Likewise, the university collects no data on "Gender Expansive" populations, which differs from the methodology of this survey. One choice that concerns me is the choice to groups transgender men and women in the "Gender Expansive" category, despite their identities falling within the traditional gender binary. "Gender Expansive" for non-binary or genderfluid is in line with current definitions of those terms, many readers would likely assume that transgender men and women would be grouped with their cisgender counterparts. 

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(population)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <-population |> 
  sample_n(100)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |>
  summarize(mean_100 = mean(age), sd_100 = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
population |>
  summarize(mean_total = mean(age), sd_total = sd(age))

```


### Task 6: Compare the sample and population statistics

**A** The mean of the sample was slightly lower, and the standard deviation slightly higher, the original population that the sample pulled from. The two are comparable, and the sample is similar enough to be used as representative of the wider population.

```{r}
sample_50 <-population |> 
  sample_n(50)
sample_50 |>
  summarize(mean_50 = mean(age), sd_50 = sd(age))

sample_25 <-population |> 
  sample_n(25)
sample_25 |>
  summarize(mean_25 = mean(age), sd_25 = sd(age))

```

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**: When the sample sizes get smaller, the results because less similar to those from the total population. They become less reliable samples of the total population, and less likely to be used as representative of all 1,000 individuals. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
