---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv(url("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv"))
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE The categories describe various types of crime committed within a 12-month period in Prince George's County. Some of the categories, like breaking and entering, are non-violent and have less importance to the public than other, more violent categories, such as shootings and assaults. Accident numbers are newsworthy because of their widespread impact on the county's population, but the lack of specificity within the data makes reporting on the category difficult. Accident numbers aren't seperated by road type, vehicle type or degree of injury, which could cause misleading reporting if not handled carefully.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |>
  mutate(total_incidents = rowSums(across(2:12))) |> 
  mutate(average_incidents = total_incidents/12) 
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

ANSWER HERE Accidents are the most common form of incident, which as I stated earlier, may reveal more about PG County's categorization methodology than its road safety. Auto theft, theft, theft from auto, and accidents with impound filled out the rest of the top five most occuring incidents, which again begs questions about categorization. Do burglaries fall under residential breaking or entering or thefts? Breaking and entering other is the least common incident type, so I'd look at what specific locations fall under that section, since robbery other is relatively high in comparison. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |>
  mutate(june_change = ((Jun-average_incidents)/average_incidents*100)) |>
  mutate(july_change = ((Jul-average_incidents)/average_incidents*100)) |>
  mutate(august_change = ((Aug-average_incidents)/average_incidents*100)) 
  
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER HERE While many categories have mixed results--dipping and rising around the annual average depending on the summer month--both shootings and homicides spike as much as 50% in a given summer month above the annual average. The argument could be made that while crime as a whole isn't more likely to spike during the summer season, certain severe violent crimes see much higher numbers during June, July and August.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE 
Prince George's County saw a spike in shootings and homicides this summer, according to new crime data released Wednesday.

During the months of June, July and August, homicide incidents rose an average of 37% above the annual monthly average for this year. Shootings jumped 36% above the monthly average over the same period. While summer crime has long been a concern for City Hall, the data shows that the rise in shootings and homicides are relatively isolated. Many types of crimes, including breaking and entering, theft, and robbery, rose only slightly this summer when compared to the rest of the year, or in some cases, even fell. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
