---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv(url("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv"))
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(violent_rate_2019 = violent2019/pop2019 * 10000) |>
  mutate(property_rate_2019 = property2019/pop2019 * 10000) |>
  mutate(violent_rate_2020 = violent2020/pop2020 * 10000) |>
  mutate(property_rate_2020 = property2020/pop2020 * 10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE: Looking at our new dataframe, we can see that contrary to the national narrative, Baltimore is not the epicenter of crime, either violent or property, within the state of Maryland. That distinction belongs to Ocean City on the Eastern Shore, which isn't all that surprising. Their violent crime rate doubled in a year. Baltimore still has high rates of violent crime, but many small towns in Prince George's County rank ahead of it in the property category. Property crime rates seemed especially high in towns with less than 5000 people during 2020, up from their rates the year prior. 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>
  mutate(property_change = (property_rate_2020-property_rate_2019)/property_rate_2019*100) |>
  mutate(violent_change = (violent_rate_2020-violent_rate_2019)/violent_rate_2019*100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE: Overall, it looks like crime dropped across most cities from 2019 to 2020, with the major exception of Ocean City, where violent crime jumped 83%. That would be my story: how crime Ocean City rose during the pandemic while it sank across most of the rest of the state. The outliers, unsurprisingly, are the small towns with only a handful of crimes recorded each year, so the percent changes seems dramatic, even if crime only dropped from 1 to 0. Those aren't the story here.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE: 

In a year that saw dramatic reductions in crime across most of Maryland, violent crime nearly doubled in Ocean City between 2019 and 2020, making it the most dangerous city the state.

Baltimore, which ranked as the state's most dangerous city in 2019, saw its violent crime rate dip 13% during the first year of the pandemic. The state's next largest city, Frederick, saw its violent crime decrease by nearly 96%--a stark contrast to the trends seen in Maryland's popular beach town. Ocean City recorded 91 violent incidents in 2019, which lept to 166 the year after, even while tourism slumped and lockdown orders reduced foot traffic that summer.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
