---
title: "HW7_MandMs"
date: 10/1/2025
---


In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mm_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)
sample5 <-mm_data |> 
  sample_n(5)

sample10 <-mm_data |> 
  sample_n(10)

```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 |>
  summarize(mean_5 = mean(red), median_5 = median(red), sd_5 = sd(red))

sample10 |>
  summarize(mean_10 = mean(red), median_10 = median(red), sd_10 = sd(red))

```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
mm_data |>
  summarize(mean_total = mean(red), median_total = median(red), sd_total = sd(red))

```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
mm_data |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 5) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) - sd(red)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) + sd(red)), color = "blue", linetype = "dashed", size = 1) +
  ggtitle("Histogram of Total Red M&M Population") +
  xlab("Number of Red M&Ms") +
  ylab("Number of Instances")

sample5 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 5) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) - sd(red)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) + sd(red)), color = "blue", linetype = "dashed", size = 1) +
   ggtitle("Histogram of Red M&M Population in Sample5") +
  xlab("Number of Red M&Ms") +
  ylab("Number of Instances")


sample10 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 5) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) - sd(red)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(red) + sd(red)), color = "blue", linetype = "dashed", size = 1) +
     ggtitle("Histogram of Red M&M Population in Sample10") +
  xlab("Number of Red M&Ms") +
  ylab("Number of Instances")

```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

**A** The standard deviation is the most helpful in understanding the accuracy of the samples, because the standard deviation can demonstrate how accurately sampled data represents the spread of data from a total population.

What does this exercise tell you about the relationship between sample size and estimate accuracy?

**A** This exercise shows that the larger the sample size, the most accurate its estimates will be to the total population.

How does random sampling introduce variation into statistical estimates?

**A** Random sampling, without ensuring a representative collection of certain groups, will create variation that can fail to capture the true proportion of minority proups within datasets. 

If you repeated this process, would the samples always be identical? Why or why not?

**A** If we repeated this, the samples would almost certainly be different, since the R program scoops out a different sample each time. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
