---
title: "HW9_TestScores"
Name: Derek Willis/Rob Wells
Date: 10/21/2025
---

name: Adam Hudacek


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Did a New Reading Program Lead to Better Scores?

The superintendent recently claimed that a new reading program has improved third-grade reading scores across the school district.

Before the program, third-grade students in the district averaged 72.6 points on standardized reading tests with a standard deviation of 4.8 points.

After implementing the program for one semester, you collected scores from 12 randomly selected classrooms:
74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75

As a journalist, you need to determine: **Is there statistical evidence that reading scores have actually improved?**

## Task 1: Organize your data and initial assessment

Before you can run this codeblock, you will need to fill in a value where it says REPLACE_ME. That value can be found in the introduction.

```{r}
# Known information about reading scores before the new program
prior_mean <- 72.6  # average score
prior_sd <- 4.8     # standard deviation

# Reading scores after implementing the new program (12 classrooms)
new_scores <- c(74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75) # Replace with the actual scores

# Create a journalist-friendly dataset
score_data <- tibble(
  classroom = paste("Classroom", 1:12),
  reading_score = new_scores
)

# View the data
score_data
```

### Reflection Question 1:  (2 points)
Based on just looking at the score_data dataframe, have test scores improved? How can you tell?

Scores have improved, because each classroom has a score higher than the prior mean. 

## Task 2: Calculate key statistics  (2 points)

Like Task 1, you will need to replace values where it says REPLACE_ME before running any code.


```{r}
# Calculate statistics based on the new reading scores
new_stats <- score_data |> 
  summarise(
    mean = mean(reading_score),
    sd = sd(reading_score),
    n = n()
  )

new_stats
```

### Reflection Question 2: (3 points)
Looking at the mean and standard deviation of the new scores compared to the previous statistics, what initial observations can you make? Provide examples from the data in your answer. What questions might these statistics raise for your reporting?

Scores have risen, but they also are more closer congregated together than before, as shown by the standard deviation of 1.76, down significantly from 4.8. I'm wondering what changes were made on the educator side, and if instruction was possibly standardized across classrooms, which could explain the more unified result. 

## Task 3: Create a column chart  (1 points)

As before, replace any values marked REPLACE_ME based the instructions.


```{r}
# STUDENT TASK: Choose an appropriate fill color for the bars
my_fill_color <- "cyan" # Replace with a color name like "royalblue", "darkgreen", etc.

# Create a visualization comparing new scores to the previous average
score_data |> 
ggplot(aes(x = classroom, y = reading_score)) +
  geom_col(fill = my_fill_color, alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 2, y = prior_mean - 1, 
           label = "Previous Average (72.6)",  hjust = -0.5, vjust= 2, fontface = "bold", color = "darkred") +
  labs(
    title = "Reading Scores After New Program Implementation",
    subtitle = "Horizontal line shows previous district average of 72.6 points",
    x = NULL,
    y = "Reading Test Score",
    caption = "Source: District Assessment Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Reflection Question 3: (3 points)
Examine the chart you created, and suggest a better title based on the results of the data, not a description.

2025 reading scores leap over 2024 average following new program implementation

## Task 4: Perform a hypothesis test (2 points)

This is where we formally test the superintendent's claim that reading scores have improved. Fill in the REPLACE_ME values as needed, beginning with your hypotheses.

**Hypotheses:**
Null: The new reading program did not improve scores
Alternative: The new reading program improved scores

```{r}
# Set the significance level for your test
alpha_level <- 0.05 # Replace with the appropriate value

# Perform a one-sample t-test
# Since we want to know if scores improved (increased), we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  score_data$reading_score,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

### Reflection Question 4:  (3 points)
What does the p-value tell you, and what doesn't it tell you? How would you explain these results to a non-technical audience while maintaining accuracy?

The p-value means the result is highly statistically significant. I would explain that following the implementation of the new program, the school saw an increase and standardization of reading test scores in 2205. I would explain that the data shows a correlation between the test scores and the new program. 

## Task 5: Interpreting the results for your news story (2 points)

Let's gather all of the important stats we'll need in one place, so we can look at the prior average, the new scores and the results of the t.test, including the confidence interval. Replace any values where it says REPLACE_ME.


```{r}
# Get the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(score_data$reading_score)$conf.int

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = prior_mean,
  `New average` = mean(new_scores),
  `Improvement` = mean(new_scores) - prior_mean,
  `Percent change` = round(((mean(new_scores) - prior_mean) / prior_mean) * 100, 1),
  `p-value` = p_value,
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

# Display the key statistics
story_stats
```

## Conclusion

### Reflection Question 5: (4 points)
Based on these statistics, what would be your headline and lead paragraph for this story? Is there evidence to support the superintendent's claim?

School district sees increased reading scores after new cirriculum

Hypothetical School District had something to celebrate Wednesday morning, as the results of Maryland's annual standardized testing released to students across the state. HSD, which implemented a new reading cirriculum last year, saw its efforts pay off as average test scores jumped 4%. The new education program sent the district's average score to 75.6, above last year's 72.6.

### Reflection Question 6:  (1 point)
What metrics or outcomes beyond test scores might be important to track for assessing reading performance?

I would also look at what percent of children are reading at or above grade level.

