---
title: "Final Project"
author: "Adam Hudacek"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
```

```{r}
#Import CSV into a dataframe titled 'billboard'
billboard <- read_csv("Billboard.csv") 

#Clean column names to reduce future coding errors
billboard <- billboard |>
  clean_names()

#Examine data to determine which columns are characters and which are doubles, as this will inform which significance tests we perform later on
glimpse(billboard)

```

Overview: 
This dataframe contains every song that hit No. 1 on the Billboard Hot 100 from August 4, 1958 to January 11, 2025. For each song, there is a plethora of data points, ranging from a song's production to its genres to its critical reception. For my analysis, a song's overall critical receptions and weeks at the top spot on the charts will be most salient. What we don't have are the publications or critics whose scores and opinions form the critical consensus for each song, or information on whether those publications have changed or remained constant over time. This could impact any drawn conclusions.  

Question: 
Does a Number 1 hit song's critical reception, or "Overall Rating," impact its longevity at the top position on the Billboard charts?

Alternative Hypothesis: 
Number 1 songs with higher overall ratings are more likely to have longer runs at the top position of the Billboard chart.

Null Hypothesis: 
A Number 1 song's overall rating does not impact its longevity on the top position of the Billboard chart.

```{r}
#For this question, we'll need to examine the "Overall Rating" and "Weeks at Number One" columns. 

#Calculate the Min, Max, Mean, Median and Standard Deviation of "Overall Rating"
summary(billboard$overall_rating)

#Calculate the Min, Max, Mean, Median and Standard Deviation of Weeks at Number One
summary(billboard$weeks_at_number_one)
```

```{r}

#Create a Scatterplot to determine if there is a correlation between "Weeks at Number One" and "Overall Rating"
billboard |>
  ggplot(aes(x = `overall_rating`, y = `weeks_at_number_one`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Average Critical Rating", y = "# of Weeks at Number One", title = "Scatterplot of Average Rating vs Weeks at Number One")
```

```{r}
#Run a Linear Regression Model to determine statistical significance between "Weeks at Number One" and "Overall Rating"
billboard |>
lm_result <- lm(`overall_rating` ~ `weeks_at_number_one`, data = billboard)
summary(lm_result)
```

```{r}
#Run a one-way ANOVA test to determine statistical significance between "Weeks at Number One" and "Overall Rating"
billboard |>
anova_result <- aov(`overall_rating` ~ `weeks_at_number_one`, data = billboard)
summary(anova_result)

```
Data Analysis: 
In both in the Linear Regression Model and the ANOVA test, the relationship between a song's overall critical reception and the number of weeks it spends at number one on the Billboard charts is statistically significant, as the p-value generated by those tests is about 0.00000141, far below the significance threshold of 0.05. However, the adjusted r-squared value of about 2% tells us that the variables have a small relationship, and other factors impact a song's longevity at the top spot on the chart. This conclusion is supported by the the scatterplot I generated from the data, which shows a weak positive correlation between the two variables.

I conclude that I can accept the alternative hypothesis and reject the null hypothesis, as our significance tests clearly revealed a statistical relationship between a song's overall critical score and its length of stay atop the Billboard chart. 

