---
title: "Regression and Beers"
author: "Wells"
date: "2025-12-02"
output: html_document
---

This is another exercise to break down how a simple linear regression works and how to use it in real life.
The topic: beer.

This following exercise comes from Diez, D., Cetinkaya-Rundel, M., & Barr, C. (2024). OpenIntro Statistics (4th ed.). http://www.openintro.org/redirect.php?go=os&referrer=os4_pdf, p 335


Many people believe that gender, weight, drinking habits, and many other factors are much more important in predicting blood alcohol content (BAC) than simply considering the number of drinks a person consumed. Here we examine data from sixteen student volunteers at Ohio
State University who each drank a randomly assigned number of cans of beer. These students were evenly divided between men and women, and they differed in weight and drinking habits. Thirty minutes later, a police officer measured their blood alcohol content (BAC) in grams of alcohol per deciliter of blood.



Load libraries
```{r}
library(tidyverse)
```


Load data
```{r}
bac_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/activities/bac%20regression/bac_data.csv")

bac_data

#Student #1 drank 5 beers and has a blood alcohol level of 0.13, etc
```


# Linear Regression
```{r}
# Run the regression
model <- lm(bac ~ beers, data = bac_data)

# Display regression results
cat("\n\nRegression Results:\n")
print(summary(model))
```

 Plot the data and the model line
```{r}
# Create regression equation text
eq_text <- sprintf("BAC = %.4f + %.4f * beers", 
                   coef(model)[1], coef(model)[2])


ggplot(bac_data, aes(x = beers, y = bac)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
  labs(
    x = "Number of Beers",
    y = "Blood Alcohol Content (BAC)",
    title = "BAC vs Number of Beers Consumed"
  ) +
  annotate("text", x = min(bac_data$beers), y = max(bac_data$bac), 
           label = eq_text, hjust = 0, vjust = 1, size = 4) +
  theme_minimal()

```

**Task 1: Interpret the results of this regression. Answer the following questions:**

a) What the predictor or independent variable? What is the response or dependent variable? 
b) What is the regression model asking and what does it tell us?
c) Do the data provide strong evidence that drinking more cans of beer is associated with an increase in blood alcohol? 
d) State the null and alternative hypotheses, report the p-value, and state your conclusion.


**Answer:**
a) The independent variable is the number of beers and the dependent variable is the blood alcohol content. 
b) The regression model is asking whether there is a statistically significant relationship between the two variables. The resulting p-value shows that there is, since it is less than 0.05. 
c) Yes, the data provides strong evidence that drinking more cans of beer is associated with an increase in blood alcohol level. 
d) The null hypothesis is that drinking any amount of beer does not affect someone's blood alcohol content. The alternative hypothesis is that drinking more beer increases someone's blood alcohol content. The p-value from our regression model was 0.0005239, well below 0.05, which means there is a statistically significant relationship between drinking beer and blood alcohol level, although the adjusted R-squared value of 0.5591 indicates that the number of beers someone drinks isn't the only predictor of their blood alcohol content.


