---
title: "Final Project"
author: "Adam Hudacek"
date: "12-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(janitor)
```

```{r}
#Import CSV into a dataframe titled 'billboard'
billboard <- read_csv("Billboard.csv") 

#Clean column names to reduce future coding errors
billboard <- billboard |>
  clean_names()

#Examine data to determine which columns are characters and which are doubles, as this will inform which significance tests we perform later on
glimpse(billboard)

```

**Overview** 
This dataframe contains every song that hit No. 1 on the Billboard Hot 100 from August 4, 1958 to January 11, 2025. For each song, there is a plethora of data points, ranging from a song's production to its genres to its critical reception. For my analysis, a song's overall critical receptions and weeks at the top spot on the charts will be most salient. What we don't have are the publications or critics whose scores and opinions form the critical consensus for each song, or information on whether those publications have changed or remained constant over time. This could impact any drawn conclusions.  

**Question** 
Does a Number 1 hit song's critical reception, or "Overall Rating," impact its longevity at the top position on the Billboard charts?

**Alternative Hypothesis**
Number 1 songs with higher overall ratings are more likely to spend at least three weeks at the top position of the Billboard chart.

**Null Hypothesis** 
A Number 1 song's overall rating does not impact its longevity on the top position of the Billboard chart.

```{r}
#For this question, we'll need to examine the "Overall Rating" and "Weeks at Number One" columns. 

#Calculate the Min, Max, Mean, Median and Standard Deviation of "Overall Rating"
summary(billboard$overall_rating)

#Calculate the Min, Max, Mean, Median and Standard Deviation of Weeks at Number One
summary(billboard$weeks_at_number_one)
```
**Analysis** 
The average number one song on the Billboard Hot 100 has an overall rating of 6, and it spends about 3 weeks at the top of the chart. Removing the outliers--like the maximum stay of 19 weeks--the median song spends exactly 2 weeks at number one. 

```{r}

#Create a Scatterplot to determine if there is a correlation between "Weeks at Number One" and "Overall Rating"
billboard |>
  ggplot(aes(x = `overall_rating`, y = `weeks_at_number_one`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Average Critical Rating", y = "# of Weeks at Number One", title = "Scatterplot of Average Rating vs Weeks at Number One")
```

**Analysis**
The scatterplot shows a weak positive correlation between a song's average critical rating and the number of weeks it spends at number one on the Billboard charts. This may indicate that the variables have some impact on each other, but also reveals that critical ratings do not entirely predict a song's stay on the Billboard Hot 100's top spot.

```{r}
#Run a Linear Regression Model to determine statistical significance between "Weeks at Number One" and "Overall Rating"
lm_result <- lm(`overall_rating` ~ `weeks_at_number_one`, data = billboard)
summary(lm_result)
```

```{r}
#Run a one-way ANOVA test to determine statistical significance between "Weeks at Number One" and "Overall Rating"
anova_result <- aov(`overall_rating` ~ `weeks_at_number_one`, data = billboard)
summary(anova_result)

```
**Data Analysis**
In both in the Linear Regression Model and the ANOVA test, the relationship between a song's overall critical reception and the number of weeks it spends at number one on the Billboard charts is statistically significant, as the p-value generated by those tests is about 0.00000141, far below the significance threshold of 0.05. However, the adjusted r-squared value of about 2% tells us that the variables have a small relationship, and other factors impact a song's longevity at the top spot on the chart. This conclusion is supported by the the scatterplot I generated from the data, which shows a weak positive correlation between the two variables. Songs with higher average critical scores appear to be more likely to spend at least 3 weeks atop the Billboard Hot 100, which is the mean number of weeks spent at number one within the data.

It is worth noting that the LRM and ANOVA are similar significance tests that both lean on the p-value to determine significance and generate identical significance codes in their outputs. Due to the fact that both variables are numerical, t-tests and z-tests are not applicable in this instance, which is why the LRM and ANOVA were chosen, despite their similarities. 

I conclude that I can accept the alternative hypothesis and reject the null hypothesis, as our significance tests clearly revealed a significant statistical relationship between a song's overall critical score and its length of stay atop the Billboard chart. Songs with higher average critical scores are more likely to spend at least 3 weeks atop the Billboard Hot 100.

**Story Pitch**
I'd pitch a story looking at the three number one songs in 2024 with the lowest average critical ratings, which all scored below 5. The three songs--CARNIVAL, I Had Some Help and Love Somebody--were authored by musicians mired in controversy, Kanye West and Morgan Wallen, yet managed to defy the historical statistical trends to spend a cumulative 8 weeks, or 15% of the year, atop the charts. Since we discovered that there is a significant statistical relationship between a song's overall critical score and its length of stay atop the Billboard chart, these recent outliers are worth examining. 

I would want to speak to the music critics cited in the data, whose scores were calculated into the critical averages, as well as music industry and listener trend experts that could explain why these songs managed to overcome the critical consensus and cultural backlash. I would also explain the data analysis I completed for this assignment, which shows that while critical averages can impact Billboard longevity, the adjusted r-squared value of about 2% demonstrates that other factors influence the weeks spent at number one, which is likely what occurred for West's and Wallen's songs. 

A potential problem is that no matter who we speak to and what our data analysis reveals, we can never be completely certain as to why these three songs defied conventional wisdom and statistical trends. However, relying on expert testimony and analysis will allow us to draw certain conclusions, although they will need to be hedged and attributed directly to those we speak to.

